Leadership across cultures and contexts: an exploratory analysis

### 2020-05-02 leadership_across_cultures_contexts.Rmd
multivariate models of all vars; heatmap of results

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=F)
library(readr)
library(knitr)
library(pander)
library(patchwork)
library(gt)
loadd(all_study_vars)
loadd(df_qual)
loadd(df_func)
loadd(plot_func_qual)
loadd(qualities_support_txt)
loadd(qualities_support_cult)
loadd(functions_support_txt)
loadd(functions_support_cult)
loadd(highstatus_plot)
loadd(plot_costs_benefits)
loadd(plot_pubdate)
loadd(m_lpca_qual)
loadd(m_lpca_func)
source("R/cheapobjects.R")
```

@@

ZG then created a measure of group structure coded as: *residential subgroup*, which includes informal groups of co-residents, social groups, age-based groups, performance groups; *kin group*, which includes lineages, phraties, and other kin-based organizations; *economic group*, which includes subsistence groups, market groups, and other groups with primarily economic goals; *military group*, which includes all groups related to inter-group conflict; *religious group*, which includes groups collected for spiritual or supernatural purpose; *political group (community)*, which includes political groups at the level of the community, i.e., the political leader can potentially interact directly with most community members such as most village headmen, or *political group (supracommunity)*, which includes political groups that encompass multiple residential communities, such as complex chiefdoms, regional political leaders, and kings or state-level leaders. ZG and KS then coded all text records for group structure.

@@

This exploratory study extracted the qualities, functions, benefits, and costs of leadership from the ethnographic record. The variables on leader qualities and functions were *not* determined prior to the study but were instead identified and coded during multiple passes through `r nrow(text_records)` texts on leadership obtained from the eHRAF (see Methods section). Our results therefore represent the broadest and most comprehensive view of the qualities of leaders (i.e., the types of people that occupy leadership roles), the functions of leaders (i.e., what leaders do), and the costs and benefits of leadership for both leaders and followers across a diverse and culturally representative sample of the ethnographic record. These results reveal universal dimensions of human leadership while also suggesting important variation by group type, subsistence type, and sex of the leader.

## Effects of culture and context on human leadership

The distribution of text records by group and subsistence type suggests leaders emerge within similar social domains among populations with varied subsistence strategies, but at variable frequencies. See Figure \@ref(fig:grouptype). Exploratory analyses identified both universal and variable dimensions of human leadership, across cultures and contexts.

@@

Exploratory modeling revealed variation in the vast majority (INCLUDE NO. or %) of our researcher coded variables (TABLEs) did not vary by group type or subsistence strategy (ajusting for effects of leader sex and continental region and the hierarchical structure of ethnographer within culture), suggesting that many of these dimensions of leadership are potnetially universal features of human social groups or at least not strongly impacted by the within or between-group measures investigated here.

@@

Leadership in residential subgroups (e.g., social groups, performance groups) was relatively common among hunter-gatherers but relatively rare among agriculturists. Leadership at the supracommunity level, in contrast, was relatively rare among hunter-gatherers but quite common among agriculturists. The only example of supracommunity leadership among hunter-gatherers was from the Andamanese, a group of populations of Negrito hunter-gatherers of the Andaman Islands by [@az02-001, p. 45], who discussed the roles of a "head chief" from a permanent residential location and their marginal influence over other, local residential leaders (see quote in Methods section).

@@

Relationships between leadership and high status, including dependencies and causality are puzzling and currently the focus of much work in evolutionary leadership studies [e.g., @van_vugt_dual_2019; @smith_leadership_2020; @cheng_why_2020; @cheng_dominance_2019; @von_rueden_dynamics_2019]. Drawing on a comparative-evolutionary perspective, incorporating findings from nonhuman social species, @van_vugt_dual_2019 emphasize that although leadership and high social status do indeed often covary, leaders are not necessarily high-status individuals, exemplified when low-ranking individuals in nonhuman species emerge as leaders of group movement or when younger and lower status men and women initiate cooperative activities (such as raids in tribal warfare or social-political movements). @cheng_dominance_2019 and @cheng_why_2020, emphasize, however, that social hierarchies form principally in response to inter-individual variation in status, which may be gained via distinct pathways. Thus, according to @cheng_why_2020, although all leaders are not necessarily high status, especially across nonhuman species, among humans high social status may be a necessary but insufficient condition for achieving and maintaing positions of leadership. In our view, social context is critical to these discussions, as is often mentioned but also under-emphasized and not well supported empirically. Processes of leadership and followership emerge across levels of social organization and individuals simultaneously maintain multiple social statuses. The "low status" male who initiates a inter-group raid on a neighboring group may not be widely respected within his community nor be a candidate for community leadership, however, he might be of exceptionally high status among his peers and high ranking within what @garfield_review_2019 [p. 71] called the "intragenerational social hierarchy."

Our results offer some insight into relationships between leadership and social status across cultures and contexts. As mentioned, high status is the most common quality of leaders (Figure \@ref(fig:varsupport)) and the second most common benefit of leadership and the loss of status the most frequent cost (Figure \@ref(fig:varsupportcb)). 
<!-- mentioned in this sample of the ethnographic record which includes discussions of chiefs, work group leaders, shamans, and age-sets, from a culturally diverse sample including foraging groups and kingdoms.  -->
This lends support to the notion of high social status often being a necessary condition of leader emergence [@cheng_why_2020]. In predicting the ethnographic texts which described leaders as high status, leaders of kin groups and leaders among hunter-gatherers and pastoralists were identified as more likely to be high status relative to some other group and subsistence types. Although discussions of female leaders were rare, the presence of female leaders were also associated with high social status (see Figure \@ref(fig:groupstatus)). Interestingly, political leaders were less likely to be described as high status, which could be related to institutionalized and overt or ascribed social distinctions. Nonetheless, these findings question the informed speculation by @smith_leadership_2020 that, "leadership may therefore be more strongly correlated with high status in large, complex organizations, such as in corporations and governments (p. 2)." However, our text analytic results, which relied on the semantic context of text records to predict the presence of high status leaders generally identified terms related to status and prestige (i.e., *respect*, *prestige*, *honor*) and possibly dominance or proscribed hierarchy (i.e., *rank*, *position*), as stronger predictors of high status leader than terms associated with leadership per se (i.e., *influence*, *authority*, *headman*), which lends some support to the distinction between status and leadership highlighted by @smith_leadership_2020 (though we caution from drawing strong interpretations from these text-analytic results, see Figure \@ref(fig:highstatusplot)). Overall, these results suggest associations between leadership and high status are ubiquitous across human societies, including among kin groups and among hunter-gatherers and that while effective leaders across a wide range of social contexts are the recipients of increased social status for their valuable services, ineffective leadership is associated with reduced social status.

@@

We performed hierarchical cluster analysis on our leader quality and leader function variables to determine if these variables formed theoretically meaningful groups. Variables were clustered together if they were positively correlated. Leadership is expected to be associated with particular suites of characteristics and functions, which potentially vary by cultural or social context, personality, or via other inter-individual variation [@cheng_two_2013; @von_rueden_roots_2014; @hogan_personality_2020; @lukaszewski_extraversion_2015; @henrich_evolution_2001; @van_vugt_dual_2019]. Much of the evolutionary theoretical literature on leadership has suggested social influence is most often achieved via one of two distinct strategies: prestige or dominance. @van_vugt_dual_2019 [p. 1], for example, "argue that there are two leadership styles: prestige style and dominance style," and @cheng_two_2013 [p. 119] in focusing on dominance and prestige as two distinct leadership strategies suggest that, "human social stratification is characterized by the co-occurrence of both strategies."

The dominance-prestige theory as formulated by @henrich_evolution_2001 significantly advanced evolutionary approaches to leadership and inspired much subsequent theoretical and empirical work. Exploratory cluster analyses reported here, however, suggest multiple features of leader qualities and leader functions across cultures and contexts, beyond dominance and prestige. Leaders likely rely a range of individual capacities, including physical, supernatural, and cognitive endowments, in order to organize group members, mediate problems, and provide prosocial services to the group, all while conforming to social and cultural pressures.

@@

The Prosocial qualities cluster comprised subclusters we labeled, "Cultural conformity" and "Prosocial competencies", the later of which included a mix of positive personal traits, such as generosity and proper behavior, as well as interpersonal skills, having a favorable personality, oratory skill, and charisma, which generally correspond to a charismatic style of leadership that has been identified in industrialized settings and widely discussed in sociology and more recently emphasized in evolutionary psychology models of leadership [@den_hartog_culture_1999; @weber_distribution_1978; @grabo_charismatic_2016; @howell_role_2005]. The larger Prosocial qualities cluster also corresponds to @henrich_big_2015's Big-man mechanism, which suggests positive assortment of followers to prosocial leaders can facilitate the evolution of cooperative social norms and prestige-based leadership. A few variables did not seem to fit prosocial theme, however, such as *Physical health*, *Physical formidable*, and *Drug consumption* (although drug use is often an important social activity).

The three most frequent leader quality variables were found in the Individual qualities cluster, which had two smaller and strongly supported subclusters: "Social, reproductive, and material success", which included traits indicative of the benefits of successful leadership, and a more general "Competencies" cluster which included qualities that are not necessarily prosocial. Leaders, probably as a consequence of their ability to provide social benefits and impose social costs, are also characterized by their material, social, and reproductive success (bottom cluster Figure \@ref(fig:cluster)A), which is consistent with much of the theoretical literature focused on associated outcomes of leadership and social hierarchy [e.g., @von_rueden_mens_2016; @bowles_emergence_2010; @glowacki_role_2013]. Again, some variables in this subcluster, such as *Ingroup member* and *Loyalty*, did not quite fit the theme. Several of these measures of success often co-occur. Among the Trobriands, horticulturalists in Papua New Guinea, @ol06-037 [p. 47] describes relationships between polygynous marriages, increased wealth, and widespread social contacts:

@@

@@

```{r FunctionRE, fig.width=10, fig.cap="Ethnographer and Culture random effects for function variables."}
plot_fun_auth_cult_re
```

```{r QualityRE, fig.width=10, fig.cap="Ethnographer and Culture random effects for quality variables."}
plot_qual_auth_cult_re
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=F)
library(readr)
library(knitr)
library(pander)
library(patchwork)
library(gt)
loadd(all_study_vars)
loadd(df_qual)
loadd(df_func)
loadd(plot_func_qual)
loadd(qualities_support_txt)
loadd(qualities_support_cult)
loadd(functions_support_txt)
loadd(functions_support_cult)
loadd(highstatus_plot)
loadd(plot_costs_benefits)
loadd(plot_pubdate)
loadd(m_lpca_qual)
loadd(m_lpca_func)
loadd(multi_allvars)
loadd(p_heatmap_subsis)
loadd(p_heatmap_region)
loadd(p_heatmap_sex)
loadd(p_heatmap_groups)
source("R/cheapobjects.R")
```

@@

ZG then created a measure of group structure coded as: *residential subgroup*, which includes informal groups of co-residents, age-based groups, performance groups; *kin group*, which includes lineages, phraties, and other kin-based organizations; *economic group*, which includes subsistence groups, market groups, and other groups with primarily economic goals; *military group*, which includes all groups related to inter-group conflict; *religious group*, which includes groups collected for spiritual or supernatural purpose; *political group (community)*, which includes political groups at the level of the community, i.e., the political leader can potentially directly interact with most community members such as most village headmen, or *political group (supracommunity)*, which includes political groups that encompass multiple residential communities, such as complex chiefdoms, regional political leaders, and kings or state-level leaders. ZG and KS then coded all text records for group structure.

@@

This exploratory study extracted the qualities, functions, benefits, and costs of leadership from the ethnographic record. The variables on leader qualities and functions were *not* determined prior to the study but were instead identified and coded during several passes through `r nrow(text_records)` texts on leadership obtained from the eHRAF (see Methods section). Our results therefore represent the broadest and most comprehensive view of the qualities of leaders (i.e., the types of people that occupy leadership roles), the functions of leaders (i.e., what leaders do), and the costs and benefits of leadership for both leaders and followers across a diverse and culturally representative sample of the ethnographic record. These results reveal universal dimensions of human leadership while also suggesting important variation by group type, subsistence type, and sex of the leader.

## The group context of human leadership

The distribution of text records by group and subsistence type suggests leaders emerge within similar social domains among populations with varied subsistence strategies, but at variable frequencies. See Figure \@ref(fig:grouptype).

@@

@@

Leadership in residential subgroups (e.g., hunting party, work group) was relatively common among hunter-gatherers but relatively rare among agriculturists. Leadership at the supracommunity level, in contrast, was relatively rare among hunter-gatherers but quite common among agriculturists. The only example of supracommunity leadership among hunter-gatherers was from the Andamanese, a group of populations of Negrito hunter-gatherers of the Andaman Islands by [@az02-001, p. 45], who discussed the roles of a "head chief" from a permanent residential location and their marginal influence over other, local residential leaders (see quote in Methods section).

@@

Relationships between leadership and high status, including dependencies and causality are puzzling and currently the focus of much work in evolutionary leadership studies [e.g., @van_vugt_dual_2019; @smith_leadership_2020; @cheng_why_2020; @cheng_dominance_2019; @von_rueden_dynamics_2019]. Drawing on a comparative-evolutionary perspective, incorporating findings from nonhuman social species, @van_vugt_dual_2019 emphasize that although leadership and high social status do indeed often covary, leaders are not necessarily high-status individuals, exemplified when low-ranking individuals in nonhuman species emerge as leaders of group movement or when younger and lower status men and women initiate cooperative activities (such as raids in tribal warfare or social movements). cheng_dominance_2019 and cheng_why_2020, emphasize, however, that social hierarchies form principally in response to inter-individual variation in status, which may be gained via distinct pathways. Thus, although all leaders are not necessarily high status, especially across nonhuman species, among humans social status may be a necessary but insufficient condition for achieving and maintaing positions of leadership. In our view, social context is critical to these discussions, as is often mentioned but also under-emphasized and not well supported empirically. Processes of leadership and followership emerge across levels of social organization and individuals simultaneously maintain multiple social statuses. The "low status" male who initiates a inter-group raid on a neighboring group may not be widely respected within his community or be a contending Big man, but he might be of exceptionally high status among his peers.

Our results offer some insight into relationships between leaders and social status across cultures and contexts. As mentioned, high status is the most common quality of leaders (Figure \@ref(fig:varsupport)) and the second most common benefit of leadership and a loss of status the most frequent cost (Figure \@ref(fig:varsupportcb)) mentioned in this sample of the ethnographic record which includes discussions of chiefs, work group leaders, shamans, and age-sets, from a culturally diverse sample including foraging groups and kingdoms. This lends support to the notion of high social status often being a necessary condition of leader emergence [@cheng_why_2020]. In predicting the ethnographic texts which described leaders as high status, leaders of kin groups and leaders among hunter-gatherers and pastoralists were identified as more likely to be high status relative to some other group and subsistence types. Although discussions of female leaders were rare, the presence of female leaders were also associated with high social status (see Figure \@ref(fig:groupstatus)). Interestingly, political leaders were less likely to be described as high status, which could be related to institutionalized and overt or ascribed social distinctions. Nonetheless, these findings question the informed speculation by @smith_leadership_2020 that, "leadership may therefore be more strongly correlated with high status in large, complex organizations, such as in corporations and governments (p. 2)." However, our text analytic results, which relied on the semantic context of text records to predict the presence of high status leaders generally identified terms related to status and prestige (i.e., *respect*, *prestige*, *honor*) and possibly dominance or proscribed hierarchy (i.e., *rank*, *position*), as stronger predictors of high status leader than terms associated with leadership per se (i.e., *influence*, *authority*, *headman*), which lends some support to the distinction between status and leadership highlighted by @smith_leadership_2020 (though we caution from drawing strong interpretations from these text-analytic results, see Figure \@ref(fig:highstatusplot)). Overall, these results suggest associations between leadership and high status are ubiquitous across human societies, including among kin groups and among hunter-gatherers and that while effective leaders across a wide range of social contexts are the recipients of increased social status for their valuable services, ineffective leadership is associated with reduced social status.

@@

We performed hierarchical cluster analysis on our leader quality and leader function variables to determine if these variables formed theoretically meaningful groups. Variables were clustered together if they were positively correlated. Leadership and high social status are expected to be associated with particular suites of characteristics and functions, which potentially vary by cultural or social context, personality, or via other inter-individual variation [@cheng_two_2013; @von_rueden_roots_2014; @hogan_personality_2020; @lukaszewski_extraversion_2015; @henrich_evolution_2001; @van_vugt_dual_2019]. Much of the evolutionary theoretical literature on leadership has suggested social influence is most often achieved via one of two distinct strategies: prestige or dominance. @van_vugt_dual_2019 [p. 1], for example, "argue that that there are two leadership styles: prestige style and dominance style," and @cheng_two_2013 [p. 119] in focusing on dominance vs. prestige as two distinct leadership strategies suggest that, "human social stratification is characterized by the co-occurrence of both strategies."

We acknowledge that the dominance-prestige theory as formulated by @henrich_evolution_2001 has significantly advanced evoluytioanry approaches to leadership, however, exploratory cluster analyses reported here suggest multiple features of leader qualities and leader functions across cultures and contexts. Leaders likely rely a range of individual capacities, including physical, supernatural, and cognitive endowments, in order to organize group members, mediate problems, and provide prosocial services to the group, all while conforming to social and cultural pressures.

@@

The Prosocial qualities cluster comprised subclusters we labeled, "Cultural conformity" and "Prosocial competencies", the later of which included a mix of positive personal traits, such as generosity and proper behavior, as well as interpersonal skills, having a favorable personality, oratory skill, and charisma, which generally correspond to a charismatic style of leadership that has been identified in industrialized settings and widely discussed in sociology and more recently emphasized in evolutionary psychology models of leadership [@den_hartog_culture_1999; @weber_distribution_1978; @grabo_charismatic_2016; @howell_role_2005]. The larger Prosocial qualities cluster also correspondes to @henrich_big_2015's Big-man mechanism, which suggests positive assortment of followers to prosocial leaders can facilitate the evolution of cooperative social norms and prestige-based leadership. A few variables did not seem to fit prosocial theme, however, such as *Physical health*, *Physical fromidable*, and *Drug consumption* (although drug use is often an important social activity).

The three most frequent leader quality variables were found in the Individual qualities cluster, which had two smaller and strongly supported subclusters: "Social, reproductive, and material success", which included traits indicative of the benefits of successful leadership, and a more general "Competencies" cluster which included qualities that are not necesssarily prosocial. Leaders, probably as a consequence of their ability to provide social benefits and impose social costs, are also characterized by their material, social, and reproductive success (bottom cluster Figure \@ref(fig:cluster)A), which is consistent with much of the theoretical literature focused on associated outcomes of leadership and social hierarchy [e.g., @von_rueden_mens_2016; @bowles_emergence_2010; @glowacki_role_2013]. Again, some variables in this subcluster, such as *Ingroup member* and *Loyalty*, did not quite fit the theme. Several of these measures of success often co-occur. Among the Trobriands, horticulturalists in Papua New Guinea, @ol06-037 [p. 47] describes relationships between polygynous marriages, increased wealth, and widespread social contacts:

@@

## Relative effects of subsistence strategy, region, leader sex, and group structure on all variables

```{r heatmapsubsis}

plot(p_heatmap_subsis)
```

```{r heatmapregion}

plot(p_heatmap_region)
```

```{r heatmapsex}

plot(p_heatmap_sex)
```

```{r heatmapgroups}

plot(p_heatmap_groups)
```

@@

```{r FunctionRE, fig.cap="Ethnographer and Culture random effects for function variables."}
plot_fun_auth_cult_re
```

```{r QualityRE, fig.cap="Ethnographer and Culture random effects for quality variables."}
plot_qual_auth_cult_re
```
